Copulas and their potential for ecology

### 2019-10-10 Paper.Rmd
added discussion paras

<!--***DAN: remianing to do:
DONE 1) Edit abstract
2) Go through the whole thing one more time
3) send it out to coauthors (done) and maybe get feedback
4)make sure to say something about tropical systems
5) maybe write to the editors saying this is coming soon?
6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.
7) change the box around the Cedar Creek data in the main fig
8) say in Discussion that you could have used non-ranks for the nonparam stats - this
is the stuff the Ecol Mongr referee talked about. Maybe this should go in Background instead?
9) maybe go back and read the Ecol Monogr reviews again
-->

@@

In addition to our model selection approach, we also used a
nonparametric approach, to provide greater 
confidence in our answers to Q1. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics, with some additional details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All our variables were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Fig. \ref{NonparamStatsFig}). 
Our statistics quantify the association
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
association in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about association in 
the right parts. The first statistic, a partial Spearman correlation
\begin{equation}
\cor_{l_b,u_b}(u,v)=\frac{\sum 
   (u_i-\mean(u)) (v_i-\mean(v))}{(n-1)\sqrt{\var(u)\var(v)}},
\end{equation}
\noindent is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds. 
Here sample means and variances are computed using all $n$ data 
points, but the
sum is over only the indices $i$ for which $u_i+v_i > 2l_b$ and 
$u_i+v_i < 2u_b$.
Larger values of the partial Spearman correlation indicate stronger positive association.
The sum of 
$\cor_{0,0.5}(u,v)$ and $\cor_{0.5,1}(u,v)$ (or some other choice of 
$\cor_{l_{b_k},u_{b_k}}(u,v)$ for bounds $l_{b_k},u_{b_k}$ that partition the
interval $(0,1)$) equals the standard Spearman correlation
as long as no points fall directly on the bounds.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar qualitative
interpretation to $\cor_{l_{b},u_{b}}$.
Our third statistic, $\Dsq_{l_b,u_b}$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong association
between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong association. These statistics are not estimators of the 
tail dependence quantities
defined previously, but rather are conceptually similar measures of associations in
the tail portions of the distributions when appropriate values of $l_b$ and $u_b$ are
selected.

@@

@@

<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->
Our results showed that the skewness of the 
spatial average of local time series is influenced 
by their tail associations. But the same logic should also apply to any collection of time
series, whether associated with locations in space or not. 
Another potential application is time series of abundances of 
all species from a single community, e.g., all
plants in a quadrat surveyed repeatedly over time. A large 
literature has focused on synchrony versus 
compensatory dynamics between such time series, and 
the influence of interspecific relationships on the 
variability of community
properties such as total biomass (e.g., @Doak1998, @Tilman1999, @Tilman2006,
@Gonzalez2009). Typically, variability of community biomass 
is measured with the coefficient of variation, but skewness
may also be of interest because it can help characterize "spikiness" 
through time. Future work on copula 
structure of interspecific relationships in communities and its implications for community 
variability is likely to be valuable, in our opinion.

>>>>

<!--***DAN: remianing to do:
DONE 1) Edit abstract
2) Go through the whole thing one more time
3) send it out to coauthors (done) and maybe get feedback
4)make sure to say something about tropical systems
DONE 5) maybe write to the editors saying this is coming soon?
6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.
DONE 7) change the box around the Cedar Creek data in the main fig
8) say in Discussion that you could have used non-ranks for the nonparam stats - this
is the stuff the Ecol Mongr referee talked about. Maybe this should go in Background instead?
9) maybe go back and read the Ecol Monogr reviews again
-->

@@

In addition to our model selection approach, we also used a
nonparametric approach, to provide greater 
confidence in our answers to Q1. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics, with some additional details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All our variables were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Fig. \ref{NonparamStatsFig}). 
Our statistics quantify the association
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
association in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about association in 
the right parts. The first statistic, a partial Spearman correlation
\begin{equation}
\cor_{l_b,u_b}(u,v)=\frac{\sum 
   (u_i-\mean(u)) (v_i-\mean(v))}{(n-1)\sqrt{\var(u)\var(v)}},\label{eq:partialspearman}
\end{equation}
\noindent is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds. 
Here sample means and variances are computed using all $n$ data 
points, but the
sum is over only the indices $i$ for which $u_i+v_i > 2l_b$ and 
$u_i+v_i < 2u_b$.
Larger values of the partial Spearman correlation indicate stronger positive association.
The sum of 
$\cor_{0,0.5}(u,v)$ and $\cor_{0.5,1}(u,v)$ (or some other choice of 
$\cor_{l_{b_k},u_{b_k}}(u,v)$ for bounds $l_{b_k},u_{b_k}$ that partition the
interval $(0,1)$) equals the standard Spearman correlation
as long as no points fall directly on the bounds.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar qualitative
interpretation to $\cor_{l_{b},u_{b}}$.
Our third statistic, $\Dsq_{l_b,u_b}$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong association
between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong association. These statistics are not estimators of the 
tail dependence quantities
defined previously, but rather are conceptually similar measures of associations in
the tail portions of the distributions when appropriate values of $l_b$ and $u_b$ are
selected.

@@

<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->
We also elaborated potential consequences of copula structure for ecological understanding. 
Our results showed that the skewness of the 
spatial average of local time series is influenced 
by their tail associations. But the same logic should also apply to any collection of time
series, whether associated with locations in space or not. 
Another potential application is time series of abundances of 
all species from a single community, e.g., all
plants in a quadrat surveyed repeatedly over time. A large 
literature has focused on synchrony versus 
compensatory dynamics between such time series, and 
the influence of interspecific relationships on the 
variability of community
properties such as total biomass (e.g., @Doak1998, @Tilman1999, @Tilman2006,
@Gonzalez2009). Typically, variability of community biomass 
is measured with the coefficient of variation, but skewness
may also be of interest because it can help characterize "spikiness" 
through time. Future work on copula 
structure of interspecific relationships in communities and its implications for community 
variability is likely to be valuable, in our opinion.

@@

<!--Alternative methods-->

<!--Partial Person correlation-->
We carried out most of our analyses on data ranks, $u_i$ and $v_i$, for good reasons 
int he Inrtoduction, Background, and methods sections and 
reviewed here; but under some circumstances it may also be appropriate to use 
techniques which parallel our nonparametric statistics but that use unranked data, 
$x_i$, $y_i$. As mentioned previously, @Genest2007 and others
recommend carrying out inferences about dependence structures (which was our
goal) using normalized ranks, stating explicitly "statistical inference concerning
dependence structures should always be based on ranks." The reason for this is
that measures of dependence which use unranked data conflate information about
the maginal distributions of variables with information about the dependence
between variables. To provide an ecological example, suppose the true population densities $p_{1,t}$
and $p_{2,t}$ of two species of fish in a lake in year $t$ are unknown,
but are assessed using catch per unit effort (CPUE), a standard approach in fisheries
science. The CPUE measurements, which we denote $c_{1,t}$ and $c_{2,t}$ may be more
or less correlated through time than the true densities would be, if they were known,
if the function $f$ relating $p_{i,t}$ and $c_{i,t}$ is nonlinear and if Pearson
correlation is used. The function $f$ modifies the marginal distributions, which modifies
measures such as Pearson correlation that conflate dependence and marginal information.
Rank-based measures such as Spearman or Kendall correlation, or any of our copula approaches,
will be unaffected by nonlinear functions such as $f$. The same difficulty will pertain in any
case for which a measurement of an ecological quantity is a nonlinear index of the true quantity
of interest. Nevertheless, if one is certain that measurements are linearly related to 
the true quantities being measured, statistics based on unranked data may be more appropriate 
under some circumstances, in part because ecological mechanisms which produce relationships 
between variables are influenced not only by the ordinal informatio in ranks but also 
by the cardinal information in the unranked data. Judgements of what statistics
are appropriate depend, of course, on the purpose of the analysis; the goal of this 
study was to study dependence in isolation, so we used ranks. If one is interested
in tail associations using cardinal data, a partial Pearson correlation can developed
straightforwardly by replacing $u$ and $v$ in equation \ref{eq:partialspearman}
by the $x$ and $y$, but still computing the sum in that equation over data points
with ranks, $u_i$ and $v_i$ constrained by the bounds $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$. This would result in a different approach to
tail associations that would conflate dependence and marginal information but 
that may also be useful for understanding ecological relationships.

<!--6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.-->
Copula methods are numerous, and go well beyond the cases we have considered
in this introductory paper; we hope our work helps inspire applications of copulas
in ecology beyond the specific tools we have employed. For instance, multivariate copulas are useful for 
studying relationships between multiple interacting variables, and have been used in finance
to model tail risk for portfolio optimization problems [@Czado2019; @Joe2011]. 
Analogous ecological applications
exist that may be amenable to the same approach, e.g., ecosystem functioning variables
such as total primary productivity or carbon flux are the sum of the contributions of
multiple species in the same way that athe value of an investment portfolio is the sum 
of the values of its constituent assets. Ecologists may likewise be interested in
managing the risk that an ecosystem functioning variable will take an extreme value. 
As another example, we assumed for this paper that univariate marginal 
distributions have continuous, strictly 
monotonic distribution functions; this was for simplicity and because such a simpler
approach was sufficient for our research questions. But count data in ecology have been analyzed
with approaches not making this assumption [@Anderson2018].
